---
title: Worksheet
---

We saw in [Intro to R part 1](https://bristol-training.github.io/intro-r-1/pages/060-data.html) how to load a dataset with `read.csv`. This was a study about [gene expression data of 42 ER- and ER+ breast cancer patients](https://ftp.ncbi.nlm.nih.gov/geo/datasets/GDS3nnn/GDS3716/soft/GDS3716.soft.gz). The code was like:

```r
breast <- read.csv("https://raw.githubusercontent.com/Bristol-Training/beginning-r/refs/heads/main/data/GDS3716.soft",
            sep="\t",
            skip=99)
```

**Using `tidyverse`**, answer the below exercises.

::: {#exercise-1 .callout-note title="Exercise 1" icon=false} 
Read the dataset. You may want to look at the documentation for [`read_delim` function](https://readr.tidyverse.org/reference/read_delim.html). Remember specifying the separator character and number of lines you want to skip.
:::

::: {#answer1 .callout-caution icon=false title='Answer 1' collapse="true"}
```{r}
library(tidyverse)

breast <- read_delim("https://raw.githubusercontent.com/Bristol-Training/beginning-r/refs/heads/main/data/GDS3716.soft",
            delim="\t",
            skip=99)
```

The function `read_delim` is quite smart and if we don't specify the delimiter character will figure it out, although it may not always work.
```r
breast <- read_delim("https://raw.githubusercontent.com/Bristol-Training/beginning-r/refs/heads/main/data/GDS3716.soft",
            skip=99)
```
:::



::: {#exercise-2 .callout-note title="Exercise 2" icon=false}
Calculate the average expresion levels for genes matching the name TP53.
:::

::: {#answer2 .callout-caution icon=false title='Answer 2' collapse="true"}

First we can filter the rows that match the gene name and remove the non-numeric columns.

```{r}
expr_tp53 <-
    breast %>%
        filter(IDENTIFIER=="TP53") %>%
        select(!c(ID_REF, IDENTIFIER)) 

expr_tp53
```

Now to calculate the mean of each one of the rows we can run
```{r}
expr_mean_tp53 <-
    breast %>%
        filter(IDENTIFIER=="TP53") %>%
        select(!c(ID_REF, IDENTIFIER)) %>%
        rowMeans()

expr_mean_tp53
```

:::



::: {#exercise-3 .callout-note title="Exercise 3" icon=false} 
Find the patient with the higher average expression levels accross the whole genome.
:::

::: {#answer3 .callout-caution icon=false title='Answer 3' collapse="true"}

```{r}
max_idx <-
    breast %>%
        select(!c(ID_REF, IDENTIFIER))  %>%
        colMeans(na.rm=TRUE) %>%
        which.max()

max_idx

```

:::

